---
title: "Lecture 3: Discrete Random Variables"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## R Preparation

### Conditional Statements

`if` statement takes a condition and evaluates the immediately following 
expressions(s) only when the condition evaluates to `TRUE`. You can use `{` to
indicate which set of expressions should be evaluated conditionally.

You can specify the expression(s) to run when the condition evaluates to `FALSE`
using `else`. `else` statement should follow immediately after the expression(s)
for the `if` statement.

If you have more than one condition to check, you can place them in order using
`else if` statement. `else` statement must be placed at the end if there is one.
The statements are evaluated in order and when a condition evaluates to `TRUE`,
R skips the remaining statements.

You can omit `else` when no action is required if all conditions are `FALSE`.


```{r}
#| eval: false
if (<condition 1>) {
  <expressions to run when condition 1 is TRUE>
} else if (<condition 2>) {
  <expressions to run when condition 1 is FALSE and condition 2 is TRUE>
} else {
  <expressions to run when FALSE>
}
<expressions run regardless of the condition>
```

> For simple conditional operations where you are assigning a single variable
conditionally, you can often use `ifelse()` and abbreviate your code. 
See `?ifelse`.

### Common distributions in R

In the previous two R lessons, we used `sample` function to simulate random choices between a finite set of objects. Because all finite sets are countable, they are examples of _discrete_ distributions. 

R also provides specific functions for common distributions. For each distribution, R provides 

+   **`dxxx`** for computing the pmf for discrete and pdf for continuous^1^ distribution;
+   **`pxxx`** for computing the cdf;
+   **`qxxx`** for computing quantiles^1^; and
+   **`rxxx`** for simulating a value from the distribution.

See R's documentation on distributions for a full list of distributions covered by default.

```{r distributions, exercise=TRUE}
?distributions
```

_^1^We will cover continuous random variables and quantiles during next lecture._


### Practice

Find the appropriate function for generating random samples from a Poisson 
distribution and complete the following code chunk so that it prints a true
statements every time you run where `Y` is sum of 3 Poisson variables with
$\lambda=5$.

```{r completerpois, exercise=TRUE, exercise.lines=15}
Y <- # add 3 randomly generated Poisson variables
print(paste("Y is ", Y))
if ( ) {
  print("Y is greater than 17.")
} else if ( ) {
  print("Y is less than or equal to 17.")
} else {
  print("Y is less than 15.")
}
```

```{r completerpois-hint}
paste("STA", "237", "H") # concatenates two or more character values together
```

```{r completerpois-solution}
Y <- sum(rpois(3, 5))
print(paste("Y is ", Y))
if (Y > 17) {
  print("Y is greater than 17.")
} else if (Y >= 10) {
  print("Y is less than or equal to 17.")
} else {
  print("Y is less than 15.")
}
```

## Simulating Rock Paper Scissors

Recall the example from class. Winners take \$3, \$2, and \$1 from the loser
when the winner wins with a rock, a paper, and scissors respectively. If it's
a tie, both players lose \$3.

Let $S$ be the random variable representing the net dollar amount you earn
after player 3 rounds with a friend. Estimate $P(S>5)$ using `1,000,000` rounds
of simulations.

```{r rps, exercise=TRUE, exercise.lines=25}
N <- 100 # start with a smaller number of simulations as you work on your code
winning_hand <- matrix(
  sample(c("Rock", "Paper", "Scissors"), N * 3, replace = TRUE),
  nrow = N
)
winner <- matrix(
  sample(c("You", "Friend", "Tie"), N * 3, replace = TRUE),
  nrow = N
)
you_earn <- numeric(N)
for (i in seq_len(N)) {
  if () { # in the the case of winning
    
  } else if () { # in the the case of losing
    
  } else {
    
  }
}
```

```{r rps-hint-1, exercise.lines=25}
N <- 100 
winning_hand <- matrix(
  sample(c("Rock", "Paper", "Scissors"), N * 3, replace = TRUE),
  nrow = N
)
winner <- matrix(
  sample(c("You", "Friend", "Tie"), N * 3, replace = TRUE),
  nrow = N
)
you_earn <- numeric(N)
for (i in seq_len(N)) {
  for (j in seq_len(3)) {
    if (winner[i, j] == "You") { # in the the case of winning
      # the amount depends on the winner's hand
    } else if (winner[i, j] == "Friend") { # in the the case of losing
      # the amount depends on the winner's hand
    } else {
      # this is when the game is tied and you automatically lose $3
      you_earn[i] <- you_earn[i] - 3
    } 
  }
}
sum(you_earn > 5) / N
```

```{r rps-hint-2, exercise.lines=25}
N <- 1000000
winning_hand <- matrix(
  sample(c("Rock", "Paper", "Scissors"), N * 3, replace = TRUE),
  nrow = N
)
winner <- matrix(
  sample(c("You", "Friend", "Tie"), N * 3, replace = TRUE),
  nrow = N
)
you_earn <- numeric(N)
for (i in seq_len(N)) {
  for (j in seq_len(3)) {
    if (winner[i, j] == "You") { # in the the case of winning
      # what happens if you omit else?
      if (winning_hand[i, j] == "Rock") {
        you_earn[i] <- you_earn[i] + 1
      }
      if (winning_hand[i, j] == "Paper") {
        you_earn[i] <- you_earn[i] + 2
      } 
      if (winning_hand[i, j] == "Scissors") {
        you_earn[i] <- you_earn[i] + 3
      }
    } else if (winner[i, j] == "Friend") { # in the the case of losing
      if (winning_hand[i, j] == "Rock") {
        you_earn[i] <- you_earn[i] - 1
      }
      if (winning_hand[i, j] == "Paper") {
        you_earn[i] <- you_earn[i] - 2
      } 
      if (winning_hand[i, j] == "Scissors") {
        you_earn[i] <- you_earn[i] - 3
      }
    } else {
      # this is when the game is tied and you automatically lose $3
      you_earn[i] <- you_earn[i] - 3
    } 
  }
}
sum(you_earn > 5) / N
```

```{r rps-hint-3, exercise.lines=25}
# you can avoid using the loop
N <- 1000000
winning_hand <- matrix(
  sample(c("Rock", "Paper", "Scissors"), N * 3, replace = TRUE),
  nrow = N
)
winner <- matrix(
  sample(c("You", "Friend", "Tie"), N * 3, replace = TRUE),
  nrow = N
)
amount <- ifelse(
  winner == "Tie",
  3, # when the game is a tie the amount is 3
  ifelse( # check for the winner's hand
    winning_hand == "Rock", 1, 
    ifelse(winning_hand == "Paper", 2, 3)
  )
)
earn_or_lose <- ifelse(winner == "You", 1, -1) # 1 for your earning, -1 for your loss
you_earn <- rowSums(amount * earn_or_lose)
sum(you_earn > 5) / N
```

> You may use `seq_len(N)` in place of  `1:N`.

<!-- data frame and plotting next -->
<!-- ## Poisson and Binomial Distributions -->

<!-- Read Example 3.48 from Devore and Berk's textbook Modern Mathematical Statistics with Applications and replicate Figure 3.8 with $\text{Bin}(30, .1)$, $\text{Bin}(100, .03)$, $\text{Bin}(300, .01)$, and $\text{Pois}(3)$. You may omit the vertical bars. -->

<!-- ```{r poisbin, exercise=TRUE, exercise.lines=20} -->
<!-- x <- 0:10 -->
<!-- df <- data.frame( -->
<!--   x = x, -->
<!--   bin30 = dbinom(x, size = 30, prob = 0.1), -->
<!--   bin100 = numeric(11), # replace appropriately -->
<!--   bin300 = numeric(11), # replace appropriately -->
<!--   pois = numeric(11) # replace appropriately -->
<!-- ) -->
<!-- ggplot(df, aes(x = x)) + -->
<!--   theme_minimal() + -->
<!--   geom_point(aes(y = bin30, pch = "Bin(30, 0.1)")) + # adding pch within mapping creates a legend based on point type -->
<!--   labs(y = "P(x)", pch = "Distribution") -->
<!-- ``` -->


<!-- ## Side: Plotting with `ggplot2` -->

<!-- ```{r plotbinom, exercise=TRUE, exercise.lines=15, warning=F} -->
<!-- library(ggplot2) # load the library -->
<!-- ggplot(binomN, aes(x = n)) + # initialize a ggplot canvas with the data -->
<!--   theme_classic() + # optional: control the overall look with predefined themes -->
<!--   geom_segment(aes(y = pmf, yend = 0, xend = n)) + -->
<!--   geom_point(aes(y = pmf)) -->
<!-- ``` -->

<!-- ```{r include=FALSE} -->
<!-- library(ggplot2) -->
<!-- binomN$cdf <- pbinom(0:10, 10, 1/4) -->
<!-- ``` -->


<!-- ```{r echo=FALSE, warning=FALSE, message=FALSE} -->
<!-- library(patchwork) -->
<!-- library(tidyverse) -->
<!-- binomN <- add_row( -->
<!--   binomN, n = -1, pmf = 0, cdf = 0, .before = 1) %>% -->
<!--   add_row(n = 11, pmf = 0, cdf = 1) -->
<!-- ppmf <- ggplot(binomN, aes(x = n)) + -->
<!--   theme_classic() + -->
<!--   geom_segment(aes(y = pmf, yend = 0, xend = n)) + -->
<!--   geom_point(aes(y = pmf)) + -->
<!--   labs(y = expression(p[N]), title = "Probability mass function") + -->
<!--   scale_x_continuous(breaks = seq(0, 10, 5)) -->
<!-- pcdf <- ggplot(binomN, aes(x = n, y = cdf)) + -->
<!--   theme_classic() + -->
<!--   geom_segment(aes(xend = lead(n), yend = cdf), stat = 'identity') + -->
<!--   geom_point(data = filter(binomN, n > -1, n < 11)) + -->
<!--   labs(y = expression(F[N]), title = "Cumulative distribution function") + -->
<!--   scale_y_continuous(limits = c(0, 1)) + -->
<!--   scale_x_continuous(limits = c(-1, 11), breaks = seq(0, 10, 5)) -->
<!-- # plot both -->
<!-- ppmf +  pcdf -->
<!-- ``` -->


## Example: Sum of a Fair Die

_(Exercise 4.5 from Dekking et al.)_

A fair die is thrown until the sum of the results of the throws exceeds 6. The random variable $X$ is the number of throws needed for this. Let $F$ be the cumulative distribution function of $X$. Determine $F(1)$, $F(2)$, and $F(7)$.

### Approach 1

> We know $X$ does not exceed $7$. We can 1) simulate 7 rolls in each iteration,
2) add individual rolls until the sum exceeds 6, and 3) `break` to stop the loop 
for adding the rolls.

```{r ex2, exercise=TRUE, exercise.lines=15}
N <- 1000000
rolls <- matrix(
  sample(1:6, N * 7, replace = TRUE),
  nrow = N
)
X <- numeric(N)
for (i in seq_len(N)) {
  for (j in 2:7) {
    s <- sum(rolls[i, (1:j)]) # sum up to j
    if (s > 6) { # check whether it exceeds 6
      X[i] <- j  # store j if it exceeds 6
      break
    }
  }
}
sum(X <= 1) / N
sum(X <= 2) / N
sum(X <= 7) / N
```

### Approach 2

> `while ()` is another type of loop similar to `for ()` except it loops _while_
the condition inside `()` is `TRUE`. It can run indefinitely if the condition
is always `TRUE`.

```{r ex1, exercise=TRUE, exercise.lines=18}
N <- 1000
X <- numeric(N)
for (i in seq_len(N)) {
  sum_of_rolls <- 0
  while (sum_of_rolls < 7) {
    # roll the die and add to the number of rolls
  }
} 
sum(X <= 1) / N
sum(X <= 2) / N
sum(X <= 7) / N
```

```{r ex1-solution, exercise=TRUE, exercise.lines=18}
N <- 1000000
X <- numeric(N)
for (i in seq_len(N)) {
  sum_of_rolls <- 0
  while (sum_of_rolls < 7) {
    sum_of_rolls <- sum_of_rolls + sample(1:6, 1)
    X[i] <- X[i] + 1
  }
} 
sum(X <= 1) / N
sum(X <= 2) / N
sum(X <= 7) / N
```


### Approach 3: Homework

Compute the values without simulation.


