---
title: "Lecture 3: Discrete Random Variables"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Discrete Distributions in R

In the previous two R lessons, we used `sample` function to simulate random choices between a finite set of objects. Because all finite sets are countable, they are examples of _discrete_ distributions. 

R also provides specific functions for common distributions. For each distribution, R provides 

+   **`dxxx`** for computing the pmf for discrete and pdf for continuous^1^ distribution;
+   **`pxxx`** for computing the cdf;
+   **`qxxx`** for computing quantiles^1^; and
+   **`rxxx`** for simulating a value from the distribution.

See R's documentation on [distributions](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/Distributions) for a full list of distributions covered by default.

_^1^We will cover continuous random variables and quantiles during next lecture._

### Example: 10 rounds of Rock Paper Scissors

$$W_i\sim\text{Ber}(1/3)\quad\text{for }i = 1,2,...,10$$
$$N=\sum_{i=1}^{10} W_i\sim\text{Bin}(10,1/3)$$

For a binomial distribution, we can use **`dbinom`**, **`pbinom`**, **`rbinom`**, and **`qbinom`**.

To compute $p_N(n)$ for $n=0,1,2,3,\dots,10$, we use **`dbinom`**.

```{r echo=TRUE}
pmfN <- dbinom(0:10, 10, 1/4)
# creating a data frame to represent data
binomN <- data.frame(n = 0:10, pmf = pmfN) 
binomN
```

We can use the pmf to compute the cdf or use **`pbinom`**.

```{r pbinom, exercise=TRUE, exercise.lines=5}

```

----------------

#### Side: Plotting with `ggplot2`

```{r plotbinom, exercise=TRUE, exercise.lines=15, warning=F}
library(ggplot2) # load the library
ggplot(binomN, aes(x = n)) + # initialize a ggplot canvas with the data
  theme_classic() + # optional: control the overall look with predefined themes
  geom_bar(aes(y = pmf), stat = 'identity', width =  0.02) +
  geom_point(aes(y = pmf)) 
```

```{r include=FALSE}
library(ggplot2)
binomN$cdf <- pbinom(0:10, 10, 1/4)
```


```{r echo=FALSE, warning=FALSE, message=FALSE}
library(patchwork)
library(tidyverse)
binomN <- add_row(
  binomN, n = -1, pmf = 0, cdf = 0, .before = 1) %>%
  add_row(n = 11, pmf = 0, cdf = 1)
ppmf <- ggplot(binomN, aes(x = n)) + 
  theme_classic() + 
  geom_bar(aes(y = pmf), stat = 'identity', width =  0.02) +
  geom_point(aes(y = pmf)) +
  labs(y = expression(p[N]), title = "Probability mass function") +
  scale_x_continuous(breaks = seq(0, 10, 5))
pcdf <- ggplot(binomN, aes(x = n, y = cdf)) + 
  theme_classic() +
  geom_segment(aes(xend = lead(n), yend = cdf), stat = 'identity') +
  geom_point(data = filter(binomN, n > -1, n < 11)) +
  labs(y = expression(F[N]), title = "Cumulative distribution function") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(-1, 11), breaks = seq(0, 10, 5))
# plot both
ppmf +  pcdf
```

----------------

## Poisson and Binomial Distributions

Read Example 3.48 from Devore and Berk's textbook Modern Mathematical Statistics with Applications and replicate Figure 3.8 with $\text{Bin}(30, .1)$, $\text{Bin}(100, .03)$, $\text{Bin}(300, .01)$, and $\text{Pois}(3)$. You may omit the vertical bars.

```{r poisbin, exercise=TRUE, exercise.lines=20}
x <- 0:10
df <- data.frame(
  x = x,
  bin30 = dbinom(x, size = 30, prob = 0.1),
  bin100 = numeric(11), # replace appropriately
  bin300 = numeric(11), # replace appropriately
  pois = numeric(11) # replace appropriately
)
ggplot(df, aes(x = x)) +
  theme_minimal() +
  geom_point(aes(y = bin30, pch = "Bin(30, 0.1)")) + # adding pch within mapping creates a legend based on point type
  labs(y = "P(x)", pch = "Distribution")
```


## Simulating a Distribution

### Example: Sum of a Fair Die

_(Exercise 4.5 from MIPS)_

A fair die is thrown until the sum of the results of the throws exceeds 6. The random variable $X$ is the number of throws needed for this. Let $F$ be the distribution function of $X$. Determine $F(1)$, $F(2)$, and $F(7)$.

#### Approach 1

```{r ex1, exercise=TRUE, exercise.lines=18}
n <- 10^3
# simulation approach 1
res1 <- replicate(
  n, {
    x <- 0; s <- 0
    while(s < 7) {
      s <- s + sample(1:6, 1)
      x <- x + 1
    }
    x
  }
)
cat('F(1)=', sum(res1 <= 1)/n)
cat('F(2)=', sum(res1 <= 2)/n)
cat('F(7)=', sum(res1 <= 7)/n)
```



#### Approach 2

+  We know $X$ takes one of $\{2,3,4,5,6,7\}$.

```{r ex2, exercise=TRUE, exercise.lines=15}
n <- 10^6
s <- matrix(
  sample(1:6, n*7, replace = TRUE),
  ncol = 7
)
apply(s, 1, sum) # applying a function over each row

```

#### Approach 3: Homework

Compute the values without simulation. 


